Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-01-31 draft_lm1.tex
completely re-written introduction

% Template for PLoS
% Version 2.0 July 2014
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% Be advised that this is merely a template 
% designed to facilitate accurate translation of manuscript content 
% into our production files. 
%
% This template contains extensive comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template 
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication and enters production, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% DO NOT ADD EXTRA PACKAGES TO THIS TEMPLATE unless absolutely necessary.
% Packages included in this template are intentionally
% limited and basic in order to reduce the possibility
% of issues during our production process.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For sideways tables, use the {rotating} package and use \begin{sidewaystable} instead of \begin{table} in the appropriate section. PLOS guidelines do not accomodate sideways figures.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment, and for any superscripts or subscripts also include the base number/text. For example, use $mathrm{mm}^2$ instead of mm$^2$ (do not use \textsuperscript command).
%
% DO NOT USE the \rm command to render mathmode characters in roman font, instead use $\mathrm{}$
% For bolding characters in mathmode, please use $\mathbf{}$ 
%
% Please add line breaks to long equations when possible in order to fit our 2-column layout. 
%
% For inline equations, please do not include punctuation within the math environment unless this is part of the equation.
%
% For spaces within the math environment please use the \; or \: commands, even within \text{} (do not use smaller spacing as this does not convert well).
%
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt]{article}
\usepackage{todonotes}
\usepackage{subfigure}
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{hyperref}
% line numbers
\usepackage{lineno}
% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
%\usepackage{rotating}
% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the attack ratio of Dengue Epidemics, under time-varying 
force of infection, from aggregated case notification data}
}
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Getulio Vargas (FGV), 
Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo 
Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

Dengue is a disease which is caused by 4 different types of viruses 
which co-circulate between the vector and human populations. Immunity is 
considered permanent after the exposure of humans to each of the viruses and 
cross immunity is considered very limited\cite{halstead_dengue_2007}.
Thus, the proportions of viral types co-circulating at any point in time is 
strongly dependent on previous incidence of the disease, which determines the 
number of susceptibles for each viral type at any point in time.

Dengue transmission is also modulated by environmental conditions, among which, 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease. So it is fair to say that these environmental fluctuations play 
a key role in determining beginning and end of epidemic periods through its 
effects on the force of infection, which cannot be taken as 
constant\cite{reiner_time-varying_2014}

The long term dynamics of dengue is also modulated by the alternation of virus 
types in circulation. Demographics also plays a role in replenishing the 
population of susceptibles.

The attack ratio (AR) of a disease is a measure of morbidity defined as the 
number of new cases divided by the population at risk.
For Dengue outbreaks and epidemics, it can be difficult to calculate the AR due 
to the lack of knowledge of the population at risk. The population at risk 
in the case of dengue, is the number of susceptibles to the circulating virus 
type(s) before a given epidemic.

In order to calculate the attack ratio, we need to determine the number of 
susceptibles to the circulating virus types right before the epidemic.
Without regular virological surveys, it is virtually impossible to determine 
the population at risk.

The attack ratio is also influenced by the reproductive number of the disease 
~\cite{bacaer_final_2009, katriel_attack_2012}, which is closely associated 
with the force of infection. Thus the incorporation of the effective 
reproduction number, \rr, as a function of time, is crucial to an accurate 
estimation of the AR.

There are other sources of uncertainty impinging on the estimation of the AR
of a dengue epidemic. One of them is that there are a large number of 
asymptomatic cases in any epidemic, which nevertheless acquire immunity.
It is estimated that for every case reported, 10-20 are not seen by health 
authorities (personal communication).
Another is under-reporting, which is a serious issue in Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that approximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to fully determine.

In this paper, we estimate the number of susceptibles before every major 
outbreak ($S_{0, year}$) in the last 18 years.
Due to the alternation of circulating serotypes, we detected fluctuating levels 
of susceptibility to the circulating virus(es).
To estimate the susceptibles we have set the 
transmissibility from the $R_t$ series derived from data~\cite{nishiura} 
(\ref{eq:effbeta}). 
Then, 
from the incidence series and the population at risk, we could then calculate 
the attack ratio for each epidemic.

To accomplish the estimation of the $S_0$, we use a simplified model of 
dengue transmission based on a single-strain Susceptible-Infectious-Removed 
(SIR) model.
The reason for the choice of a single strain model, is the lack of 
 information regarding the actual proportions of each virus in 
circulation.
However, some information is available about the predominant 
serotypes for some epidemics in the period 
of study~\cite{macedo_virological_2013}.

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of Dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

The authors have applied similar methods as used here, to estimate the number 
of susceptibles for the predominant strain of Influenza at the beginning of  
Influenza seasons in europe~\cite{pone2011}.

@@

In this section we will start by describing the data and then the method used 
to estimate the effective reproductive number, $R_t$, from the data and obtain 
its posterior distribution. 
We then proceed to describe the Susceptible-Infectious-Recovered (SIR) model 
used to represent the aggregated disease incidence and how $R_t$ can be 
integrated into the model to allow for time varying force of infection. 
Next, an approach to approximate the posterior distributions 
of the numbers of susceptible to the main circulating Dengue viruses for each 
epidemic is detailed.
Finally, we discuss how to estimate the attack ratio of each 
epidemic using the estimated susceptible fraction and the observed incidence.

\subsection*{The data set} 
The data used in this paper consists of time 
series of weekly notified cases of Dengue for the 
city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical criteria. Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013). The incidence is normalized for the parameter estimation 
procedures. The normalization was done by dividing the number of cases reported 
by the total city population at each year as given by the census.

@@

@@

\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$.
The authors are also grateful to Claudia T. 
Code\c{c}o for helpful discussions about the manuscript.

@@

An advantage of the Bayesian approach is that one can devise prior 
distributions for $\theta_t$ taking advantage of the intuitive parameterization 
and flexibility of the beta family of distributions.
Prior elicitation can also be done for \rr and the hyper-parameters directly 
plugged into the prior for $\theta_t$. 
One can, for example, choose prior mean and variance for \rr and find $a_0$ 
and $b_0$ that satisfy those conditions.
Let $m_0$ and $v_0$ be the prior expectation and variance for $R_t$. 
After some tedious algebra one finds
\begin{align}
\label{seq:elicitation}
a_0 &= \frac{m_0v_0 + m_0^3 + m_0^2}{v_0} \\
b_0 &= \frac{2v_0 + m_0^2 + m_0}{v_0}
\end{align}
If one wants only to specify $m_0$ and a coefficient of variation $c$ 
\footnote{$c = \sqrt{v_0}/ m_0$.} for $R_t$ \textit{a priori}, some less boring 
algebra gives:
\begin{align}
\label{seq:elicitationcv}
a_0 &= \frac{m_0^3c^2 + m_0^3 + m_0^2}{m_0^2c^2} \\
b_0 &= \frac{2m_0^2c^2 + m^2 + m}{m_0^2c^2}
\end{align}

This approach thus makes possible to incorporate epidemiological knowledge 
about disease biology (e.g. the magnitude of $R_0$) into the computation of \rr.
This may prove particularly important when disease counts are low and/or close 
to the detection threshold.
We provide an R script to perform the above elicitation at 
\url{https://github.com/fccoelho/paperLM1/blob/master/aux/elicit_Rt_prior.R}.

>>>>

% Template for PLoS
% Version 2.0 July 2014
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
\documentclass[10pt]{article}
\usepackage{todonotes}
\usepackage{subfigure}
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{hyperref}
% line numbers
\usepackage{lineno}
% ligatures disabled
\usepackage{microtype}
\usepackage{indentfirst}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
%\usepackage{rotating}
% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the Attack Ratio of Dengue Epidemics under Time-varying 
Force of Infection using Aggregated Notification Data}
}
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Getulio Vargas (FGV), 
Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo 
Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

Dengue is an arthropod-borne febrile disease caused by a flavivirus with four 
serotypes which causes an estimated  $50$ million infections each 
year~\cite{Guzman2010}.
In humans, immunity against  a particular serotype is  considered permanent 
after the exposure and cross immunity to there serotypes is considered short 
lived~\cite{halstead_dengue_2007}.
As a consequence, the proportions of viral serotypes co-circulating at any 
point in time are strongly dependent on previous incidence patterns of the 
disease, which determine the number of individuals susceptible to each serotype 
at any point in time.
% LM: Maybe a citation here?

Dengue transmission is also modulated by environmental conditions, among which, 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient seasonal temperature variation, dengue is 
predominantly a  summer disease.
So it is fair to say that these environmental 
fluctuations play a key role in determining beginning and end of epidemic 
periods.
This climatic influence is exerted mainly through its effects on the force of 
infection, which cannot be taken as constant~\cite{reiner_time-varying_2014} 
but rather as a seasonal (oscillating) function of time.
The long term dynamics of dengue is also modulated by the alternation of virus 
types in circulation. Demographics also plays a role in replenishing the 
population of susceptibles.

The attack ratio (AR) of a disease is a measure of morbidity defined as the 
number of new cases divided by the population at risk.
% LM: Maybe a citation here? It's ridiculous I know, but referees can be a pain.
For dengue epidemics, it can be difficult to calculate the AR due 
to the lack of knowledge of the population at risk.
The population at risk in this case is the number of susceptibles 
to the circulating virus type(s) before a given epidemic. 
Thus, in order to calculate the attack ratio, we need to determine the number 
of susceptibles to the circulating virus types right before the epidemic, which 
is virtually impossible without regular virological surveys.

The attack ratio is also influenced by the reproductive number of the disease 
~\cite{bacaer_final_2009, katriel_attack_2012}, which is closely associated 
with the force of infection.
Thus the incorporation of the effective reproduction number, \rr, as a function 
of time, is crucial to an accurate estimation of the AR of seasonal diseases 
like dengue and Influenza.

It is also important to consider other sources of uncertainty impinging on the 
estimation of the AR of a dengue epidemic. 
For instance, in any given epidemic there is a large number of 
mild and asymptomatic cases, which nevertheless acquire immunity.
Depending on surveillance structure it is estimated that for every case 
reported, up to 10-20 are not seen by health authorities (personal 
communication).
%LM: Best to name who gave that. Oswaldo perhaps?
Another source of uncertainty is under-reporting, which is a serious issue in
the health care systems of many underdeveloped countries such as Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, meaning that approximately 37\% of the 
suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways 
which 
are not easy to fully determine.
%LM: Since this paragraph pretty much deals with possible limitations we may 
% want to move it to the discussion. see #14

Other methods for estimating the  number of susceptibles while 
accommodating time-varying force of infection have been proposed 
before, for measles~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}, a disease that shows remarkable seasonality.
These methods try to reconstruct the entire 
series of infectious and susceptibles from case data using deterministic models 
and generally work well for measles because there is a one-to-one 
relationship between exposure and immunity, since measles is caused by a 
single-strain pathogen.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergence of 
new serotypes, an intrinsically stochastic events, can drastically change the 
pool of susceptibles, throwing off any sequential estimation based on the 
incidence dynamics.

In this paper, we propose a new approach to estimate the number (fraction) of 
susceptibles using a simplified model of dengue transmission based on a 
single-strain Susceptible-Infectious-Removed (SIR) model with time-varying 
infection rate.
In order to bypass the limitations of not knowing the serotype-specific 
seroprevalence and the exact behavior of the force of infection through time, we
propose to inform the time-varying transmissibility using the \rr series 
derived from the notification data~\cite{nishiura}.
We extend a Bayesian framework previously used to estimate the number 
of susceptibles in Influenza epidemics in Europe~\cite{pone2011} to include 
time-varying force of infection and derive a probability distribution  for 
\rr to accommodate uncertainty in the estimates. 
Then, from the incidence series and the population at risk, we  calculate the 
attack ratio for each epidemic.
We apply our method to estimate $S_0$ before every major dengue epidemic in the 
city of Rio de Janeiro, Brazil in the last 18 years.

@@

In this section we will start by describing the data and then the method used 
to estimate the effective reproductive number, $R_t$, from the data and obtain 
its posterior distribution. 
We then proceed to describe the Susceptible-Infectious-Recovered (SIR) model 
used to represent the aggregated disease incidence and how $R_t$ can be 
integrated into the model to allow for time varying force of infection. 
Next, an approach to approximate the posterior distributions 
of the numbers of susceptible to the main circulating dengue viruses for each 
epidemic is detailed.
Finally, we discuss how to estimate the attack ratio of each 
epidemic using the estimated susceptible fraction and the observed incidence.

\subsection*{Data} 
The data used in this paper consists of time 
series of weekly notified cases of dengue for the 
city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical criteria. Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013). The incidence is normalized for the parameter estimation 
procedures. The normalization was done by dividing the number of cases reported 
by the total city population at each year as given by the census.

@@

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

The reason for the choice of a single strain model, is the lack of 
information regarding the actual proportions of each virus in 
circulation.
However, some information is available about the predominant 
serotypes for some epidemics in the period 
of study~\cite{macedo_virological_2013}.

@@

\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for \rr.
The authors are also grateful to Claudia T. 
Code\c{c}o for helpful discussions about the manuscript.

@@

An advantage of the Bayesian approach is that one can devise prior 
distributions for $\theta_t$ taking advantage of the intuitive parameterization 
and flexibility of the beta family of distributions.
Prior elicitation can also be done for \rr and the hyper-parameters directly 
plugged into the prior for $\theta_t$. 
One can, for example, choose prior mean and variance for \rr and find $a_0$ 
and $b_0$ that satisfy those conditions.
Let $m_0$ and $v_0$ be the prior expectation and variance for $R_t$. 
After some tedious algebra one finds
\begin{align}
\label{seq:elicitation}
a_0 &= \frac{m_0v_0 + m_0^3 + m_0^2}{v_0} \\
b_0 &= \frac{2v_0 + m_0^2 + m_0}{v_0}
\end{align}
If one wants only to specify $m_0$ and the coefficient of variation $c = 
\sqrt{v_0}/ m_0$ for $R_t$ \textit{a priori}, some less boring 
algebra gives:
\begin{align}
\label{seq:elicitationcv}
a_0 &= \frac{m_0^3c^2 + m_0^3 + m_0^2}{m_0^2c^2} \\
b_0 &= \frac{2m_0^2c^2 + m^2 + m}{m_0^2c^2}
\end{align}

This approach thus makes it possible to incorporate epidemiological knowledge 
about disease Biology (e.g. the magnitude of $R_0$) into the computation of \rr.
This may prove particularly important when disease counts are low and/or close 
to the detection threshold.
We provide an R script to perform the above elicitation at 
\url{https://github.com/fccoelho/paperLM1/blob/master/R/elicit_Rt_prior.R}.
